IFA Comparison of Noncognitive Skills \\ BCS and MCS

### 2018-03-28 codes/R/cohorts_nc.Rnw
minor change to text

\subsubsection{Model notation}
Before any meaningful comparison across cohorts can be made, measurement invariance (MI) needs to be established. Examination of the invariance properties of a measurement system is aimed at assessing whether the latent constructs in each group are measured in a comparable way, both in terms of scale and location. In the case of continuous measures, MI analysis proceeds in a relatively straightforward way \citep{Vanderberg2000a}; MI properties in systems of categorical outcomes are instead more problematic, due to the more complex nature of identification \citep{Millsap2004, Wu2016}.

I set up a (confirmatory) item factor analysis (IFA) model for the full list of 11 common items. Given the results of the exploratory analysis -- where each measure separates neatly on one of the two factors -- I assume a \emph{dedicated} factor structure \citep{Carneiro2003}, where each measure is restricted to load on one of the two factors only.\footnote{This structure is commonly assumed in economic research to enable a more straightforward interpretation of the latent constructs in question \citep{ACFMR,Heckman2012a}. It can also be denominated \emph{congeneric} factor stucture \citep{Millsap2004}.}

@@

\subsubsection{Invariance analysis}
There are infinite ways in which the above system can be parameterised and a scale obtained -- the well-known issue of factor indeterminacy. A configural model is defined as a minimal set of parameter restrictions that ensure identification of the system across groups, without restricting the model further. Widely used parameterisations for the configural model are:
\bnm[label=\Alph*:]
\item Delta parameterisation {[WE{\textDelta}]} \citep{Wu2016}

@@

\item Theta parameterisation {[WE{\textTheta}]} \citep{Wu2016}

@@

\item Anchored parameterisation {[MT]} \citep{Millsap2004}
  \bi
  \item For all groups, normalise a reference loading to 1 for each factor
  \item Set invariant across groups one threshold per item (e.g. $\tau_{0,Ai} = \tau_{0,Bi}$), and an additional threshold in the reference items above
  \item In the first group: $\bm{\kappa_A}=\bm{0}$, $\text{diag}(\bm{\Sigma}_A)=\bm{I}$
  \item Set all intercepts $\bm{\nu}$ to zero
  \ei
\enm
The first two parameterisations (WE{\textDelta} and WE{\textTheta}) work by normalising the mean and variance of factors to the same constants in both groups, and they leave all loadings and intercepts to be freely estimated. The MT parameterisation instead proceeds by first identifying one group, and then imposing cross-group equality constraints to identify other groups \citep{Wu2016}.

To test measurement invariance, one needs to estimate nested models that place increasing restrictions on the item parameters across groups $c$. In the case of continuous measures, the usual approach is to restrict one set of parameters at a time. As an example, starting from the configural model, one would restrict all loadings to be equal across groups; if the fit of this restricted model is not worse than the configural model, \emph{metric invariance} is estabilished \citep{Vanderberg2000a}. The hierarchy of the nested models usually proceeds by testing loadings first, and then intercepts (metric and scalar invariance). Metric and scalar invariance are the minimum requirements to be able to meaningfully compare means and variances of the latent factors across groups.

A substantially similar approach is proposed for categorical measures by \cite{Millsap2004}. However, new identification results in \cite{Wu2016} indicate that, in the case of categorical measures, invariance properties cannot be examined by simply restricting one set of parameters at a time. This is because the identification conditions used in the configural baseline model, while being minimally restrictive on their own, become restrictive once certain additional restrictions are imposed. In light of this, they propose sets of restrictions that identify structures of different invariance levels. They find that some restrictions cannot be tested alone against the configural model, because they give rise to statistically equivalent models. This is true of loading invariance, and also of threshold invariance in the case when the number of categories of each ordinal item is 3 or less. Furthermore, they suggest that comparison of both latent means and variances requires invariance in loadings, thresholds, and intercepts.

Following the ordering in \cite{Wu2016}, I estimate the following increasingly restrictive models:
\bi
\item \textbf{Configural} model -- see WE{\textDelta} above
\item \textbf{Threshold invariance} model
  \bi
  \item $\tau_{1,Ai} = \tau_{1,Bi}$ for $i=1,\dots,11$, $\tau_{2,Ai} = \tau_{2,Bi}$ for $i=\{1,2,3,4,7,8,9,10\}$
  \item For all groups: $$\text{diag}(\bm{\Phi}) = \bm{I}, \quad \bm{\kappa}=\bm{0}$$
  \item For the reference group: $$\bm{\nu}_A=\bm{0}, \quad \; \text{diag}(\bm{\Sigma}_A)=\bm{I}$$
  \item In our case, the number of categories in the outcomes is 3, thus this model is statistically equivalent to the configural model
  \ei
\item \textbf{Threshold and Loading invariance} model
  \bi
  \item $\tau_{1,Ai} = \tau_{1,Bi}$ for $i=1,\dots,11$, $\tau_{2,Ai} = \tau_{2,Bi}$ for $i=\{1,2,3,4,7,8,9,10\}$
  \item $\lambda_{Ai} = \lambda_{Bi}$ for $i=1,\dots,11$
  \item For all groups: $$\bm{\kappa}=\bm{0}$$
  \item For the reference group: $$\bm{\nu}_A=\bm{0}, \quad \; \text{diag}(\bm{\Sigma}_A)=\bm{I}, \quad \text{diag}(\bm{\Phi}_A) = \bm{I}$$
  \item This set of restriction allows comparison of latent factor variances
  \ei
\item \textbf{Threshold, Loading, and Intercept invariance} model
  \bi
  \item $\tau_{1,Ai} = \tau_{1,Bi}$ for $i=1,\dots,11$, $\tau_{2,Ai} = \tau_{2,Bi}$ for $i=\{1,2,3,4,7,8,9,10\}$
  \item $\lambda_{Ai} = \lambda_{Bi}$ for $i=1,\dots,11$
  \item For all groups: $$\bm{\nu}=\bm{0}$$
  \item For the reference group: $$\bm{\kappa}_A=\bm{0}, \quad \; \text{diag}(\bm{\Sigma}_A)=\bm{I}, \quad \text{diag}(\bm{\Phi}_A) = \bm{I}$$
  \item This set of restriction allows comparison of latent factor means and variances
  \ei
\ei

I estimate each of the models listed above by Weighted Least Squares.\footnote{All estimates are computed using the \texttt{lavaan} package in \textbf{\textsf{R}} \citep{Rosseel2012}. Parameters of the model are estimated by mean- and variance adjusted weighted least squares (WLSMV) (see \citealp{Mplus2017}). Point estimates for model parameters are obtained starting from the polychoric correlations, using diagonally weighted least squares (DWLS), and the full weight matrix is used to compute robust standard errors and test statistics. Robust WLS has proved in simulation studies to be moderately robust to small violations of the normality assumption in the latent underlying measures \citep{Flora2011}, and generally outperforms maximum likelihood in large samples \citep{Beauducel2009,Li2016}.} The fit of the increasingly restrictive models presented above is compared in \autoref{tab:fit}. For all models, a chi-squared difference would point to a lack of measurement invariance. However, this test is know to overestimate invariance in large samples. Alternative fit indices should thus be used when comparing nested models. \cite{Cheung2002} suggest relying on Comparative Fit Index (CFI, \citealp{Bentler1990}), McDonald Non-Centrality Index (MFI, \citealp{McDonald1989}), and Gamma Hat index \citep{Steiger1989} for the case of ordered measures. Commonly accepted thresholds for rejecting the restricted model are $\Delta CFI < - 0.01$, $\Delta MFI < - 0.02$, and $\Delta Gamma hat < - 0.001$.

The overall fit of the configural model is satisfactory according to widely accepted rules of thumb \citep{Maccallum1996}.\footnote{{\color{red} BETTER CITATIONS FOR OVERALL FIT}.} As discussed above, when items have less than four categories, the model with restricted thresholds is statistically equivalent to the configural model.\footnote{\cite{Meade2008}, using the results from a simulation study, suggests stricter thresholds that should apply in a variety of conditions. For CFI, a single cutoff values of $.002$ is proposed, while cutoffs for MFI depend on the problem's characteristics; in our case (2 factors, 11 items), they suggest $.0066$. \cite{Sass2014} however cast some doubts of the generalisability of these cutoffs to WLSMV estimators.}

@@

\clearpage
\bibliographystyle{chicago}
\bibliography{/Users/giacomomason/Documents/BibTeX/Papers-cohorts_nc}
%\bibliography{/Users/giacomomason/Dropbox/Zotero/cohorts.bib}

>>>>

\subsubsection{Model notation}
Before any meaningful comparison across cohorts can be made, measurement invariance (MI) needs to be established. Examination of the invariance properties of a measurement system is aimed at assessing whether the latent constructs in each group are measured in a comparable way, both in terms of scale and location. In the case of continuous measures, MI analysis proceeds in a relatively straightforward way \citep{Vanderberg2000}; MI properties in systems of categorical outcomes are instead more problematic, due to the more complex nature of identification \citep{Millsap2004, Wu2016a}.

I set up a (confirmatory) item factor analysis (IFA) model for the full list of 11 common items. Given the results of the exploratory analysis -- where each measure separates neatly on one of the two factors -- I assume a \emph{dedicated} factor structure, where each measure is restricted to load on one of the two factors only.\footnote{This structure is commonly assumed in economic research to enable a more straightforward interpretation of the latent constructs in question \citep{Attanasio2017c,Heckman2013}. It can also be denominated \emph{congeneric} factor stucture \citep{Millsap2004}.}

@@

\subsubsection{Invariance analysis}
There are infinite ways in which the above system can be parameterised and a scale obtained -- the well-known issue of factor indeterminacy. A configural model is defined as a minimal set of parameter restrictions that ensure identification of the system across groups, without restricting the model further. Widely used parameterisations for the configural model are:
\bnm[label=\Alph*:]
\item Delta parameterisation {[WE{\textDelta}]} \citep{Wu2016a}

@@

\item Theta parameterisation {[WE{\textTheta}]} \citep{Wu2016a}

@@

\item Anchored parameterisation {[MT]} \citep{Millsap2004}
  \bi
  \item For all groups, normalise a reference loading to 1 for each factor
  \item Set invariant across groups one threshold per item (e.g. $\tau_{0,Ai} = \tau_{0,Bi}$), and an additional threshold in the reference items above
  \item In the first group: $\bm{\kappa_A}=\bm{0}$, $\text{diag}(\bm{\Sigma}_A)=\bm{I}$
  \item Set all intercepts $\bm{\nu}$ to zero
  \ei
\enm
The first two parameterisations (WE{\textDelta} and WE{\textTheta}) work by normalising the mean and variance of factors to the same constants in both groups, and they leave all loadings and intercepts to be freely estimated. The MT parameterisation instead proceeds by first identifying one group, and then imposing cross-group equality constraints to identify other groups \citep{Wu2016a}.

To test measurement invariance, one needs to estimate nested models that place increasing restrictions on the item parameters across groups $c$. In the case of continuous measures, the usual approach is to restrict one set of parameters at a time. As an example, starting from the configural model, one would restrict all loadings to be equal across groups; if the fit of this restricted model is not worse than the configural model, \emph{metric invariance} is estabilished \citep{Vanderberg2000}. The hierarchy of the nested models usually proceeds by testing loadings first, and then intercepts (metric and scalar invariance). Metric and scalar invariance are the minimum requirements to be able to meaningfully compare means and variances of the latent factors across groups.

A substantially similar approach is proposed for categorical measures by \cite{Millsap2004}. However, new identification results in \cite{Wu2016a} indicate that, in the case of categorical measures, invariance properties cannot be examined by simply restricting one set of parameters at a time. This is because the identification conditions used in the configural baseline model, while being minimally restrictive on their own, become restrictive once certain additional restrictions are imposed. In light of this, they propose sets of restrictions that identify structures of different invariance levels. They find that some restrictions cannot be tested alone against the configural model, because they give rise to statistically equivalent models. This is true of loading invariance, and also of threshold invariance in the case when the number of categories of each ordinal item is 3 or less. Furthermore, they suggest that comparison of both latent means and variances requires invariance in loadings, thresholds, and intercepts.

Following the ordering in \cite{Wu2016a}, I estimate the following increasingly restrictive models:
\bi
\item \textbf{Configural} model -- see WE{\textDelta} above
\item \textbf{Threshold invariance} model
  \bi
  \item $\tau_{1,Ai} = \tau_{1,Bi}$ for $i=1,\dots,11$, $\tau_{2,Ai} = \tau_{2,Bi}$ for $i=\{1,2,3,4,7,8,9,10\}$
  \item For all groups: $$\text{diag}(\bm{\Phi}) = \bm{I}, \quad \bm{\kappa}=\bm{0}$$
  \item For the reference group: $$\bm{\nu}_A=\bm{0}, \quad \; \text{diag}(\bm{\Sigma}_A)=\bm{I}$$
  \item In our case, the number of categories in the outcomes is 3, thus this model is statistically equivalent to the configural model
  \ei
\item \textbf{Threshold and Loading invariance} model
  \bi
  \item $\tau_{1,Ai} = \tau_{1,Bi}$ for $i=1,\dots,11$, $\tau_{2,Ai} = \tau_{2,Bi}$ for $i=\{1,2,3,4,7,8,9,10\}$
  \item $\lambda_{Ai} = \lambda_{Bi}$ for $i=1,\dots,11$
  \item For all groups: $$\bm{\kappa}=\bm{0}$$
  \item For the reference group: $$\bm{\nu}_A=\bm{0}, \quad \; \text{diag}(\bm{\Sigma}_A)=\bm{I}, \quad \text{diag}(\bm{\Phi}_A) = \bm{I}$$
  \item This set of restriction allows comparison of latent factor variances
  \ei
\item \textbf{Threshold, Loading, and Intercept invariance} model
  \bi
  \item $\tau_{1,Ai} = \tau_{1,Bi}$ for $i=1,\dots,11$, $\tau_{2,Ai} = \tau_{2,Bi}$ for $i=\{1,2,3,4,7,8,9,10\}$
  \item $\lambda_{Ai} = \lambda_{Bi}$ for $i=1,\dots,11$
  \item For all groups: $$\bm{\nu}=\bm{0}$$
  \item For the reference group: $$\bm{\kappa}_A=\bm{0}, \quad \; \text{diag}(\bm{\Sigma}_A)=\bm{I}, \quad \text{diag}(\bm{\Phi}_A) = \bm{I}$$
  \item This set of restriction allows comparison of latent factor means and variances
  \ei
\ei

I estimate each of the models listed above by Weighted Least Squares.\footnote{All estimates are computed using the \texttt{lavaan} package in \textbf{\textsf{R}} \citep{Rosseel2012}. Parameters of the model are estimated by mean- and variance adjusted weighted least squares (WLSMV) (see \citealp{Muthen2017a}). Point estimates for model parameters are obtained starting from the polychoric correlations, using diagonally weighted least squares (DWLS), and the full weight matrix is used to compute robust standard errors and test statistics. Robust WLS has proved in simulation studies to be moderately robust to small violations of the normality assumption in the latent underlying measures \citep{Flora2004}, and generally outperforms maximum likelihood in large samples \citep{Beauducel2006,Li2016}.} The fit of the increasingly restrictive models presented above is compared in \autoref{tab:fit}. For all models, a chi-squared difference would point to a lack of measurement invariance. However, this test is know to overestimate invariance in large samples. Alternative fit indices should thus be used when comparing nested models. \cite{Cheung2002} suggest relying on Comparative Fit Index (CFI, \citealp{Bentler1990}), McDonald Non-Centrality Index (MFI, \citealp{McDonald1989}), and Gamma Hat index \citep{Steiger1989} for the case of ordered measures. Commonly accepted thresholds for rejecting the restricted model are $\Delta CFI < - 0.01$, $\Delta MFI < - 0.02$, and $\Delta Gamma hat < - 0.001$.

The overall fit of the configural model is satisfactory according to widely accepted rules of thumb \citep{Maccallum1996}.\footnote{{\color{red} BETTER CITATIONS FOR OVERALL FIT}.} As discussed above, when items have less than four categories, the model with restricted thresholds is statistically equivalent to the configural model.\footnote{\cite{Meade2008}, using the results from a simulation study, suggests stricter thresholds that should apply in a variety of conditions. For CFI, a single cutoff values of $.002$ is proposed, while cutoffs for MFI depend on the problem's characteristics; in our case (2 factors, 11 items), they suggest $.0066$. \cite{Sass2014a} however cast some doubts of the generalisability of these cutoffs to WLSMV estimators.}

@@

\clearpage
\bibliographystyle{chicago}
%\bibliography{/Users/giacomomason/Documents/BibTeX/Papers-cohorts_nc}
\bibliography{/Users/giacomomason/Dropbox/Zotero/cohorts.bib}
